---
title: "Nice Ride Project Data Wrangling"
author: "Tony Tushar Jr"
date: "2/2/2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

#Load packages
```{r}
library(dplyr)
library(tidyr)
library(lubridate)
library(readr)
library(weathermetrics)
library(timeDate)
library(ggplot2)
library(ruler)
```

#*Load and view 2016 trip history data*
```{r}
Nice_ride_trip_history_2016_season <- read_csv("Nice_ride_trip_history_2016_season.csv", 
    col_types = cols(`End date` = col_datetime(format = "%m/%d/%Y %H:%M"), 
        `Start date` = col_datetime(format = "%m/%d/%Y %H:%M")))
head(Nice_ride_trip_history_2016_season)
```

#Several NA values at end of dataset, omit
```{r}
summary(Nice_ride_trip_history_2016_season)
Nice_ride_trip_history_2016_season <- na.omit(Nice_ride_trip_history_2016_season)
tail(Nice_ride_trip_history_2016_season)
summary(Nice_ride_trip_history_2016_season)
```

#Round to nearest hour for start and end times
```{r}
Nice_ride_trip_history_2016_season$`Start date` <- round_date(Nice_ride_trip_history_2016_season$`Start date`, "hour")
Nice_ride_trip_history_2016_season$`End date` <- round_date(Nice_ride_trip_history_2016_season$`End date`, "hour")
```

#Rename columns
```{r}
Nice_ride_trip_history_2016_season <- Nice_ride_trip_history_2016_season %>% rename(Start_Date = `Start date`) %>% rename(Start_Station = `Start station`) %>% rename(Start_Station_Number = `Start station number`) %>% rename(End_Date = `End date`) %>% rename(End_Station = `End station`) %>% rename(End_Station_Number = `End station number`) %>% rename(Total_Duration_Seconds = `Total duration (seconds)`) %>% rename(Account_Type = `Account type`)
Trips_2016 <- Nice_ride_trip_history_2016_season
```

#*Load and view bike 2016 station data*
```{r}
Nice_Ride_2016_Station_Locations <- read_csv("Nice_Ride_2016_Station_Locations.csv")
head(Nice_Ride_2016_Station_Locations)
```

#Column name adjust for consistency and drop X6 variable
```{r}
Nice_Ride_2016_Station_Locations <- Nice_Ride_2016_Station_Locations %>% select(-X6) %>%  rename(Total_Docks='Nb Docks')
Stations_2016 <- Nice_Ride_2016_Station_Locations
```
----
#*Load and view 2017 trip history data*
```{r}
Nice_ride_trip_history_2017_season <- read_csv("Nice_ride_trip_history_2017_season.csv", 
    col_types = cols(`End date` = col_datetime(format = "%m/%d/%Y %H:%M"), 
        `Start date` = col_datetime(format = "%m/%d/%Y %H:%M")))
View(Nice_ride_trip_history_2017_season)
```

#Round to nearest hour for start and end times
```{r}
Nice_ride_trip_history_2017_season$`Start date` <- round_date(Nice_ride_trip_history_2017_season$`Start date`, "hour")
Nice_ride_trip_history_2017_season$`End date` <- round_date(Nice_ride_trip_history_2017_season$`End date`, "hour")
```

#Rename columns
```{r}
Nice_ride_trip_history_2017_season <- Nice_ride_trip_history_2017_season %>% rename(Start_Date = `Start date`) %>% rename(Start_Station = `Start station`) %>% rename(Start_Station_Number = `Start station number`) %>% rename(End_Date = `End date`) %>% rename(End_Station = `End station`) %>% rename(End_Station_Number = `End station number`) %>% rename(Account_Type = `Account type`) %>% rename(Total_Duration_Seconds = `Total duration (Seconds)`) 
Trips_2017 <- Nice_ride_trip_history_2017_season
```

#*Load and view bike 2017 station data*
```{r}
Nice_Ride_2017_Station_Locations <- read_csv("Nice_Ride_2017_Station_Locations.csv")
head(Nice_Ride_2017_Station_Locations)
```

#Column name adjust 
```{r}
Nice_Ride_2017_Station_Locations <- Nice_Ride_2017_Station_Locations %>% rename(Total_Docks=`Total docks`)
Stations_2017 <- Nice_Ride_2017_Station_Locations
```

#Join Trips_2016 and Stations_2016 based on Station Number
```{r}
#Start Stations
Rides_2016 <- full_join(Trips_2016, Stations_2016, by = c("Start_Station_Number" = "Terminal"))
Rides_2016 <- Rides_2016 %>% select(-Station)
Rides_2016 <- Rides_2016 %>% select(Start_Date:Start_Station, -Start_Station_Number, Latitude:Total_Docks, End_Date:Account_Type)
Rides_2016 <- Rides_2016 %>% rename(Start_Latitude = Latitude) %>% rename(Start_Longitude=Longitude) %>% rename(Start_Docks=Total_Docks)
#End Stations
Rides_2016 <- full_join(Rides_2016, Stations_2016, by = c("End_Station_Number" = "Terminal"))
Rides_2016 <- Rides_2016 %>% select(-Station, -End_Station_Number) 
Rides_2016 <- Rides_2016 %>% rename(End_Latitude=Latitude) %>% rename(End_Longitude=Longitude) %>% rename(Total_End_Docks=Total_Docks) 
Rides_2016 <- Rides_2016 %>% select(Start_Date:End_Station, End_Latitude:Total_End_Docks, Total_Duration_Seconds:Account_Type)
```

#Join Trips_2017 and Stations_2017 based on Station Number
```{r}
#Start Stations
Rides_2017 <- full_join(Trips_2017, Stations_2017, by = c("Start_Station_Number" = "Number"))
Rides_2017 <- Rides_2017 %>% select(-Start_Station_Number, -Name)
Rides_2017 <- Rides_2017 %>% select(Start_Date:Start_Station, Latitude:Total_Docks, End_Date:Total_Duration_Seconds)
Rides_2017 <- Rides_2017 %>% rename(Start_Latitude = Latitude) %>% rename(Start_Longitude=Longitude) %>% rename(Start_Docks=Total_Docks)
#End Stations
Rides_2017 <- full_join(Rides_2017, Stations_2017, by = c("End_Station_Number" = "Number"))
Rides_2017 <- Rides_2017 %>% select(-Name, -End_Station_Number) 
Rides_2017 <- Rides_2017 %>% rename(End_Latitude=Latitude) %>% rename(End_Longitude=Longitude) %>% rename(Total_End_Docks=Total_Docks) 
Rides_2017 <- Rides_2017 %>% select(Start_Date:End_Station, End_Latitude:Total_End_Docks, Total_Duration_Seconds:Account_Type)
```

#Bind 2016 and 2017 Rides
```{r}
Rides_1617 <- rbind(Rides_2016, Rides_2017)
rm(list=setdiff(ls(), "Rides_1617"))
```

#Separate start/end dates and times, add start day of week variable
```{r}
Rides_1617 <- Rides_1617 %>% mutate(Start_Year = year(Start_Date))
Rides_1617 <- Rides_1617 %>% mutate(Start_Month = month(Start_Date))
Rides_1617 <- Rides_1617 %>% mutate(Start_Day = day(Start_Date))
Rides_1617 <- Rides_1617 %>% mutate(Start_Hour = hour(Start_Date))
Rides_1617 <- Rides_1617 %>% mutate(End_Year = year(End_Date))
Rides_1617 <- Rides_1617 %>% mutate(End_Month = month(End_Date))
Rides_1617 <- Rides_1617 %>% mutate(End_Day = day(End_Date))
Rides_1617 <- Rides_1617 %>% mutate(End_Hour = hour(End_Date))
Rides_1617 <- Rides_1617 %>% mutate(Start_DoWeek = wday(Start_Date, label = TRUE))
Rides_1617 <- Rides_1617 %>% rename(End_Docks=Total_End_Docks)
```

#Create new Year-Month-Day variable for join with weather data
```{r}
Rides_1617$Start_Join <- as.Date(with(Rides_1617, paste(Start_Year, Start_Month, Start_Day, sep="-", format = "%Y-%m-%d")))
View(Rides_1617)
```

#*Load and view 2016 weather data*
```{r}
X16MSPWeatherDaily <- read_csv("16MSPWeatherDaily.csv")
head(X16MSPWeatherDaily)
dim(X16MSPWeatherDaily)
```

#Select relevant variables
```{r}
Weather_2016 <- X16MSPWeatherDaily %>% select(DATE, AWND, PRCP, SNOW, SNWD, TAVG, WT01:WT03, WT05, WT08)
View(Weather_2016)
```

#Rename variables
```{r}
Weather_2016 <- Weather_2016 %>% rename(Date = DATE, Avg_Wind = AWND, Precip = PRCP, Snow = SNOW, Snow_Depth = SNWD, Avg_Temp = TAVG, Fog = WT01, Heavy_Fog = WT02, Thunder = WT03, Hail = WT05, Haze = WT08)
```

#Replace NA's with 0 for binary variables
```{r}
Weather_2016 <- Weather_2016 %>% replace_na(list(Fog = 0, Heavy_Fog = 0, Thunder = 0, Hail = 0, Haze = 0))
```
--
#*Load and view 2017 weather data*
```{r}
X17MSPWeatherDaily <- read_csv("17MSPWeatherDaily.csv")
head(X17MSPWeatherDaily)
dim(X17MSPWeatherDaily)
```

#Select relevant variables
```{r}
Weather_2017 <- X17MSPWeatherDaily %>% select(DATE, AWND, PRCP, SNOW, SNWD, TAVG, WT01:WT03, WT05, WT08)
View(Weather_2017)
```

#Rename variables
```{r}
Weather_2017 <- Weather_2017 %>% rename(Date = DATE, Avg_Wind = AWND, Precip = PRCP, Snow = SNOW, Snow_Depth = SNWD, Avg_Temp = TAVG, Fog = WT01, Heavy_Fog = WT02, Thunder = WT03, Hail = WT05, Haze = WT08)
```

#Replace NA's with 0 for binary variables
```{r}
Weather_2017 <- Weather_2017 %>% replace_na(list(Fog = 0, Heavy_Fog = 0, Thunder = 0, Hail = 0, Haze = 0))
```

#Bind 2016 and 2017 daily weather
```{r}
Weather_1617 <- rbind(Weather_2016, Weather_2017)
View(Weather_1617)
dim(Weather_1617)
```

#Join Rides and Weather data
```{r}
Nice_Ride_1617 <- full_join(Rides_1617, Weather_1617, by = c("Start_Join" = "Date"))
head(Nice_Ride_1617)
```

#Add weekday/weekend variable for start dates
```{r}
Nice_Ride_1617 <- Nice_Ride_1617 %>% mutate(Weekend = ifelse(isWeekend(Start_Date, wday = 1:5), "1", "0"))
```

#Rearrange variables in dataset
```{r}
Nice_Ride_1617 <- Nice_Ride_1617 %>% select(Start_DoWeek, Weekend, Start_Year:Start_Hour, Start_Station:Start_Docks,  End_Year:End_Hour, End_Station:Account_Type, Avg_Wind:Haze, -Start_Date, -Start_Join, -End_Date, -Start_Hour, -End_Hour)
head(Nice_Ride_1617)
```

#Arrange table based on start month, date, and hour
```{r}
Nice_Ride_1617 <- Nice_Ride_1617 %>% arrange(Start_Year, Start_Month, Start_Day)
```

#Calculate trip length in minutes
```{r}
Nice_Ride_1617 <- Nice_Ride_1617 %>% mutate(Total_DurationMin = Total_Duration_Seconds/60)
```

#Calculate trip distance in kilometers, converted to miles
```{r}
earth.dist <- function (long1, lat1, long2, lat2)
{
rad <- pi/180
a1 <- lat1 * rad
a2 <- long1 * rad
b1 <- lat2 * rad
b2 <- long2 * rad
dlon <- b2 - a2
dlat <- b1 - a1
a <- (sin(dlat/2))^2 + cos(a1) * cos(b1) * (sin(dlon/2))^2
c <- 2 * atan2(sqrt(a), sqrt(1 - a))
R <- 6378.145
d <- R * c
return(d)
}

Nice_Ride_1617 <- Nice_Ride_1617 %>%  mutate(Trip_DistanceKM = earth.dist(Start_Longitude, Start_Latitude, End_Longitude, End_Latitude))

Nice_Ride_1617 <- Nice_Ride_1617 %>% mutate(Trip_DistanceMiles = Trip_DistanceKM/0.621371)
```

#One last rearrange of variables
```{r}
Nice_Ride_1617 <- Nice_Ride_1617 %>% select(1:17, 29, 31, 18:28, -30)
```

#Replace two off values in Account_Type variable
```{r}
table(Nice_Ride_1617$Account_Type)
Nice_Ride_1617$Account_Type <- gsub("Inconnu", "Casual", Nice_Ride_1617$Account_Type)
table(Nice_Ride_1617$Account_Type)
```

#Save tidy data
```{r}
setwd("~/Nice Ride MN 2017/Nice-Ride-MN-2017")
write_csv(Nice_Ride_1617, "Nice_Ride_1617.csv")
```






